Starcode: an exact algorithm for sequence clustering

### 2014-07-03 doc/document.tex
Edited the scalability in the manuscript.

\subsection{Lookup search}
If the Levenshtein distance of two sequences of length $L$ is
less than or equal to $\tau$, there must exist a perfect match between
them of at least $k =\lfloor L / (\tau+1)\rfloor$ nucleotides. We take
advantage of this property to build a lookup table containing all such
$k$-mers of the sequences already present in the trie. When a new
sequence is queried, all its subsequences are extracted and queried
against the lookup table. If none of the subsequences are found, the
minimum distance between the query and all the sequences in the trie
is necessarily greater than $\tau$, and the trie search can therefore
be omitted. This lookup is particulary useful for long query sequences
(see \ref{scalability}), where $k$ is large and the probability
of an exact match between subsequences is small unless they are
actually related, i.e. they belong to the same cluster. With this
method, the absence of match is detected at the computational cost of
several table lookups, compared to the higher cost of measuring the
Levenshtein distance between long sequences.

@@

\subsection{Scalability}
\label{scalability}
In Figure~\ref{fig:X} we present an evaluation of the performance and
scalability of starcode as a function of many critical parameters. The
scalability tests were performed using an **artificial set of random
sequences**. The sequence length and the number of input sequences
varies depending on the test. The standard configuration consists of a
set of **X** sequences of length **Y** running on **Z** threads and a
cluster distance of **D**. To test the scalability of starcode as a
function of a single parameter, only the selected value is modified
whereas the others are kept constant.

In Figure~\ref{fig:X}a we show the running time as a function of the
selected Levenshtein distance for the clusters. The evolution with the
increasing distance shows the common exponential trend. Note, however,
that for distances below 3 the running time remains constant. This is
due to the fact that the average sequence distance of the input
sequences lies above this threshold. Consequently, the constant part
is clearly dependent on the diversity of the input set and when the
average distance between sequences is reached, the running time scales
exponentially.

In Figure~\ref{fig:X}b we measure the running time of starcode as a
function of the number of input sequences $n$. Both axes are in
logarithmic scale, drawing an evident linear evolution. Based on this
result, we conclude that the running time of starcode is approximately
$O(n^{1.7})$.

To prove the effectiveness of the parallel processing, we ran starcode
over the same dataset using different number of threads. As shown in
Figure~\ref{fig:X}c, the algorithm is fully parallel as the relative
performance increases linearly up to 4 threads. The curve bending
produced thereafter is an artifact of having insufficient memory
bandwidth to satisfy the increased demand of memory accesses and it is
due to hardware limitations.

Finally, in Figure~\ref{fig:X}d we investigate the execution time
with varying sequence lengths. It increases initially due to the
ineffectiveness of the lookup table and the limited diversity of short
sequences, i.e. short sequences are less diverse, yielding a bigger
clustering rate. At a certain length, the lookup search algorithm
starts to be effective and boosts the performance since most of the
queries are resolved without the need for trie-search.

\begin{figure}[!tpb]%figurex
\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
\caption{Scalability. a. Running time as a function of the clustering
  distance. b. Logarithm of the running time versus the logarithm of
  the number of sequences to be clustered. c. Relative performance increase for
  different number of parallel threads. d. Running time versus length
  of the input sequences.
}\label{fig:X}
\end{figure}

@@

We generated 1 million random 40-mers that were duplicated 47 times.
For each 40-mer, we inserted 3 addional mutant sequences to the pool
by sampling again 3 nucleotides at random positions. The probability
that such a mutant sequence is within a distance 3 of any other
40-mer is of the order of $10^{-12}$ and can be neglected. Each
cluster thus consists of 50 sequences, of which 3 have a Levenshtein
distance 3 from the canonical representant.

All software were set to run on a 6-core single-processor Intel Xeon
E5-2620 system with **64~GB** of DDR3-RAM at 1333~Mhz. The execution
options for each software are summarized in
Table~\ref{tab:soft-options}. Denoted as \emph{sequential R} is the
sequential algorithm of \cite{pmid23953119} which is implemented in R
and relies on the \texttt{stringdist} function of the
\texttt{stringdist} package.

@@

Using **12** cores, starcode clustered the shuffled 50 million sequences
in 5 minutes and 17 seconds without a single error (the output
consisted of 1 million clusters of size 50).  The running times
of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
identified 2,362,352 and 2,739,954 clusters respectively. The results
are summarized in Figure~\ref{fig:benchmark}. Both algorithms rely on
heuristics, explaining why the number of clusters is off target. These
results clearly demonstrate that starcode outperforms other clustering
algorithms both in speed and accuracy with random input sequences.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics{benchmark.pdf}}
\caption{Benchmark of starcode against sequence clustering algorithms
and short read mappers. \textbf{a} Running time compared to CD-HIT and
UCLUST on the dataset described in the text. \textbf{b} Error rate
compared to CD-HIT and UCLUST, expressed as the percentage of clusters
failing to be merged. \textbf{c} Running time compared to Bowtie2, BWA
and GEM on this dataset. The running time is decomposed in extraction
of unique sequences (purple), Burrows-Wheeler indexing (yellow), and
query proper (white). The extraction was performed with the Linux sort
command, so the time is the same in every case. \textbf{d} Error rate
compared to Bowtie2, BWA and GEM, expressed as the proportion of matches
with a single hit. This is an underestimate of the true error rate.}
\label{fig:benchmark}
\end{figure}

@@

\subsection{Identifying enriched sequence motifs}
Sequence motifs are thought to play an important role in DNA metabolism. Key
regulators, such as transcription factors, nucleosomes and non coding RNAs have
sequence preferences targeting them to the sites where they act. Identifying
those sequences is a way to pinpoint the regulators and the mechanisms they
are involved in. However, the sequence motifs are not strictly identical
at different sites, hence they are better identified by inexact matching.
This problem becomes computationally difficult for long motifs (above 12-13
nucleotides) because of the combinatorial scaling. But as motifs become
longer, the problem of identifying abundant inexact matches becomes similar
to barcode clustering.

We set up a test based on the meningitis-causing agent
\textit{Neisseria meningitidis}. The genome of this bacterium is interspersed
with a frequent 12 bp sequence known as DNA uptake sequence \citep{pmid10673000}. We
extracted the 12-mers from both orientations of  the 2.19 Mb genome, yielding
4.39 million 12-mers, consisting of 2.77 million unique sequences. Clustering
the 12-mers with starcode within a Levenshtein distance of 2 took less
than 45 seconds with 12 cores. We identified the known DNA uptake sequence of
\textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant 12-mer.
This result testifies to the fact that starcode can be used to identify biologically
relevant motifs in bacterial genomes.

To test starcode on another application, we used the RNA-protein
interaction data produced by RNAcompete \citep{pmid19561594}. The mammalian
splicing factor SRSF1 is known to bind RNA GA-rich motifs, but there is
some disagreement about the motif that it recognizes \citep{pmid23562324}.
For each replicate, we replaced the RNAcompete signals by their rank and
extracted the 10-mers from the microarray probes. The 10-mers were given
a score equal to the rank of the probe they belong, and enriched motifs
were found using the sphere clustering of starcode with max Levenshtein
distance 2. The score of the most enriched 10-mer is thus the sum of the
ranks of all 10-mers within this distance. Clustering the 6.3 million
extracted 10-mers with 12 cores took about 20 seconds for each replicate.
The most enriched 10-mers were AGGACACGGA, AGGACACGGA, AGGACGGAGG,
AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except for the last replicate, the
motifs consist of AGGAC and GGA, with a spacer of variable length. This
suggests that the binding of SRSF1 to RNA is flexible, which explains the
poor agreement between the motifs derived from 6-mers or 7-mers.

>>>>

\subsection{Lookup search}
If the Levenshtein distance of two sequences of length $L$ is
less than or equal to $\tau$, there must exist a perfect match between
them of at least $k =\lfloor L / (\tau+1)\rfloor$ nucleotides. We take
advantage of this property to build a lookup table containing all such
$k$-mers of the sequences already present in the trie. When a new
sequence is queried, all its subsequences are extracted and queried
against the lookup table. If none of the subsequences are found, the
minimum distance between the query and all the sequences in the trie
is necessarily greater than $\tau$, and the trie search can therefore
be omitted. This lookup is particulary useful for long query sequences
(see \ref{performance}), where $k$ is large and the probability
of an exact match between subsequences is small unless they are
actually related, i.e. they belong to the same cluster. With this
method, the absence of match is detected at the computational cost of
several table lookups, compared to the higher cost of measuring the
Levenshtein distance between long sequences.

@@

\subsection{Performance}
\label{performance}
We measured basic performance and scalability metrics of starcode on a
dataset of pseudo random sequences (Figure~\ref{fig:perf}). The standard
configuration consists of a set of 1,000,000 sequences of length 40 running
on 1 thread and with a maximum Levenshtein distance of 3. To test the
scalability as a function of a single parameter, only the parameter under
study was modified whereas the others were kept constant.

Figure~\ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic scale
the trend is a straight line with slope 1.7, which suggests that the
running time complexity of starcode is approximately $O(n^{1.7})$.
Figure~\ref{fig:perf}b shows that the running time gros exponentially
as a function of the maximum Levenshtein distance used for clustering.
This is a common feature of exact algorithms based on trie search. As
this parameter increases, the poucet search bails out at a deeper
average depth in a trie that fans out exponentially.
The running time first increases as a function of the sequence length,
but then plummets and stays low (Figure~\ref{fig:perf}c).
Beyond a certain length, the lookup search algorithm starts to be
efficient, and most of the queries are resolved without the need for
trie-search.
Finally, in we show the scalability of starcode with increasing
number of treahds in Figure~\ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 8 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is that there is insufficient memory bandwidth to satisfy
the increased demand of memory accesses and  due to hardware limitations.

\begin{figure}[!tpb]%figurex
\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
\caption{Scalability. a. Logarithm of the running time versus the
logarithm of the number of sequences to be clustered. b. Running time as
a function of the clustering distance. c. Running time versus length of
the input sequences. d. Relative performance increase for different number
of parallel threads. 
}\label{fig:perf}
\end{figure}

@@

We generated 1 million random 40-mers duplicated 47 times. For each
40-mer, we also inserted 3 addional mutant sequences to the pool
by sampling again 3 nucleotides at random positions. The probability
that such a mutant sequence is within a distance 3 of any other
40-mer is of the order of $10^{-12}$ and can be neglected. Each
cluster thus consists of 50 sequences, of which 3 have a Levenshtein
distance 3 from the centroid of the cluster (the canonical
representant).

All software were set to run on a 6-core single-processor Intel Xeon
E5-2620 system with 50~GB of DDR3-RAM at 1333~Mhz. The execution
options for each software are summarized in
Table~\ref{tab:soft-options}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%             WHERE TO PUT THIS NOW??                               %
%Denoted as \emph{sequential R} is the
%sequential algorithm of \cite{pmid23953119} which is implemented in R
%and relies on the \texttt{stringdist} function of the
%\texttt{stringdist} package.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Using 12 cores, starcode clustered the shuffled 50 million sequences
in 5 minutes and 3 seconds without a single error (the output
consisted of 1 million clusters of size 50). There is no obvious
way to compare exact algorithms with heuristic algorithms such as
CD-HIT and USEARCH. We decided to allocate the heuristics 6 minutes
on the same dataset ($\approx$ 20\% additional time) and use the
parameters giving the best accurary under this constraint.
However, we could not find a combination of
parameters such that either of them could terminate in less than
6 minutes. To simplify the task, we extracted the unique sequences
from the input file, bringing it down to 4,000,000 and followed
the same rationale. We could not find any set of parameters such
that USEARCH could process this input in less than 6 minutes,
but CD-HIT could perform the task when run as shown in
Table~\ref{tab:soft-options}.

we decided to 
allocating them 6 minutes with the same computer resources 
USEARCH could not be
run on the same input because
The running times
of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
identified 2,362,352 and 2,739,954 clusters respectively. The results
are summarized in Figure~\ref{fig:benchmark}. Both algorithms rely on
heuristics, explaining why the number of clusters is off target. These
results clearly demonstrate that starcode outperforms other clustering
algorithms both in speed and accuracy with random input sequences.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics{benchmark.pdf}}
\caption{Benchmark of starcode against sequence clustering algorithms
and short read mappers. \textbf{a} Running time compared to CD-HIT and
USEARCH on the dataset described in the text. \textbf{b} Error rate
compared to CD-HIT and USEARCH, expressed as the percentage of clusters
failing to be merged. \textbf{c} Running time compared to Bowtie2, BWA
and GEM on this dataset. The running time is decomposed in extraction
of unique sequences (purple), Burrows-Wheeler indexing (yellow), and
query proper (white). The extraction was performed with the Linux sort
command, so the time is the same in every case. \textbf{d} Error rate
compared to Bowtie2, BWA and GEM, expressed as the proportion of matches
with a single hit. This is an underestimate of the true error rate.}
\label{fig:benchmark}
\end{figure}

@@

\subsection{Identifying enriched sequence motifs}
Sequence motifs are thought to play an important role in DNA metabolism. Key
regulators, such as transcription factors, nucleosomes and non coding RNAs have
sequence preferences targeting them to the sites where they act. Identifying
those sequences is a way to pinpoint the regulators and the mechanisms they
are involved in. However, the sequence motifs are not strictly identical
at different sites, hence they are better identified by inexact matching.
This problem becomes computationally difficult for long motifs (above 12-13
nucleotides) because of the combinatorial scaling. But as motifs become
longer, the problem of identifying abundant inexact matches becomes similar
to barcode clustering. We reasoned that starcode could also be used for the
task of identifying biologically meaningful sequence motifs.

We set up a test based on the meningitis-causing agent
\textit{Neisseria meningitidis}. The genome of this bacterium is
interspersed with a frequent 12 bp sequence known as DNA uptake
sequence \citep{pmid10673000}. We extracted the 12-mers from both
orientations of  the 2.19 Mb genome, yielding 4.39 million 12-mers,
consisting of 2.77 million unique sequences. Clustering the 12-mers
with starcode within a Levenshtein distance of 2 took less than 45
seconds with 12 cores. We identified the known DNA uptake sequence
of \textit{Neisseria meningitidis} (ATGCCGTCTGAA) as the most abundant
12-mer, with 1466 exact and 2096 inexact hits. This result testifies
to the fact that starcode can be used to identify biologically
relevant motifs in bacterial genomes.

To test starcode on another application, we used the RNA-protein
interaction data produced by RNAcompete \citep{pmid19561594}. The mammalian
splicing factor SRSF1 is known to bind RNA GA-rich motifs, but there is
some disagreement about the motif that it recognizes \citep{pmid23562324}.
For each SRSF1 replicate of the RNAcompete dataset,
we replaced the RNAcompete signals by their rank and
extracted the 10-mers from the microarray probes. The 10-mers were given
a score equal to the rank of the probe they belong, and enriched motifs
were found using the sphere clustering of starcode with maximum
Levenshtein distance 2. The score of the most enriched 10-mer is thus the
sum of the ranks of all 10-mers within this distance. Clustering the 6.3
million extracted 10-mers with 12 cores took about 20 seconds for each
replicate.  The most enriched 10-mers were AGGACACGGA, AGGACACGGA,
AGGACGGAGG, AGGACGGAGG, AGGACACGGA and AGGATACAGG. Except for the last
replicate, the motifs consist of AGGAC and GGA, with a spacer of variable
length. This suggests that the binding of SRSF1 to RNA may be flexible,
which would explain the disagreement between the motifs derived from 6-mers
or 7-mers.
